---
title: "Teacher Turnover in Wisconsin"
author: "Michael Chirico"
date: "March 17, 2017"
output:
  pdf_document:
    keep_tex: yes
    includes:
      in_header: turnover_paper_header.tex
abstract: 'Given the consistently-affirmed importance of teacher quality to student success, understanding teacher churn is crucial to formulating and evaluating teacher labor market policy. This paper replicates the analysis of @hanushek over a longer and more recent time period in Wisconsin and confirms all of its major findings, namely that while pay is a inter-district pay differentials are a significant determinant of turnover, school quality measures are much better predictors of all three types of churn -- within and between school districts and out of local public schools.'
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'hide')
knitr::read_chunk('turnover_data_cleaner.R')
library(xtable)
options(xtable.comment = FALSE)
options(xtable.include.rownames = FALSE)
library(data.table)
library(texreg)
```

```{r start_read, cache = TRUE, include = FALSE}
```

# Introduction

# Literature Review

Because the potential policy implications of turnover in the teaching profession (from human capital and equity/distributional perspectives both) are far-reaching and polypartisan, the literature on turnover-related topics in education is extensive. As relates to this paper, there are five broad (and often overlapping) categories of inquiry: the relationship between turnover and wages, which has tended to focus on “opportunity wages” outside of the field of education; the relationship between turnover, school demographics, and other nonpecuniary benefits, which has tended to focus on distributional inequalities--whether teachers with certain characteristics are more or less likely to be teaching certain disadvantaged groups; the relationship between turnover and teacher quality as measured by student performance, usually value added; collective bargaining agreements in education, focusing by and large on the implications (or lack thereof) of seniority-preferential clauses; and the recent phenomenon of specific retention incentives, the provisioning of wage bonuses to teachers willing to teach in high-needs schools.

One of the earliest papers attempting to rigorously investigate turnover was a panel study of teachers in Michigan by @murnane, who used college degree field wages outside of education as opportunity wages, finding the expected lower exit rate for teachers with higher wages in teaching relative to the authors' defined alternative. @dolton use panel data on university graduates in the United Kingdom to estimate a competing risks model of the decision to leave teaching entirely, finding results in line with @murnane. Returning to panel studies in the US, @loeb use PUMS data to get an idea of teacher relative wages in many states and find that dropout rates fall when teacher relative wages are high. @stinebrickner also uses panel data (this time NLS-72) to track both teachers and non-teachers, focusing in particular on young teachers who leave the profession for long stints, and finds that the best predictor of female exit is recent childbearing, which is an important consideration for all work related to teacher turnover because such a high percentage (76 nationwide) of teachers are female. Lastly, @hanushek focuses on teachers in Texas and emphasizes that the characteristics of students are much stronger factors in predicting teacher exit than are wages (while also affirming the statistical significance of pay).

While wages have been found consistently to have some measurable effect on teacher turnover, it is impossible to explain within-district migration (which constitutes a large portion of switching--as much as 50%) through wage-only channels because contracts are fixed at the district level. As such, another strand of literature has chosen to focus on the nonpecuniary aspects of the decision to take a teaching job--school environment/rapport, student enthusiasm, neighborhood characteristics, etc.--usually by directing attention to a single district so that any wage-based considerations are stifled, as is the case for @boyd and @engel. @boyd track early-career teachers in New York City as they quit or transfer out of the city, and most importantly finds that commuting time is an important, often overlooked aspect of location preference. @engel leverages a unique data set from Chicago Public School job fairs which affords them a rather strong measure of teachers' demand for vacancies, neutralizing the influence of school administration's behavior on turnover (through poor match selection or other means). The authors contribute evidence that the school's neighborhood (perhaps due to ambient crime or other reputational effects good and bad) is a better predictor of teachers' preference than distance from home, going somewhat against the grain of @boyd. @scafidi examine statewide data from Georgia, but ignore wage effects, choosing instead to focus on disentangling the contributions of low student achievement and minority status to turnover; they find that minority status is the more salient associate of teacher exit.

The key element missing from all of the above studies is perhaps the most important consideration in the issue of teacher turnover--teacher quality. None of the studies above have student-teacher matched data, and so are unable to directly associate student outcomes with any given teacher. If, with respect to any measure of quality you would like, we find that transitioning teachers are identical to their replacements, the issue of teacher turnover is not, in fact, much of an issue--it leans closer to hot air and wasted ink. Thus, the recent trend in the literature to incorporate measures of teacher quality (in large part made possible by a trend towards administrative records allowing students to be linked to teachers and tracked over time) in considerations of teacher turnover has made big strides in addressing the most policy-relevant questions to be asked. The most common and widely accepted measure of teacher quality is value added[^VA] (in its various guises), and the literature has begun to incorporate such measures into studies of teacher turnover. @hanushek2010 considers value added as a measure of teacher productivity, and ask if common results of labor search theory (namely that turnover falls with tenure and that turnover is negatively associated with match-specific productivity) continue to hold in the education labor market. In fact, the authors find that the teachers most likely to switch schools are those with low measured match quality, and especially that those who leave teaching entirely are those with the lowest match quality. The results are more pronounced for schools with high proportions of low-SES students, which has strong policy implications, as it appears the best teachers in high needs schools are the least likely to change jobs. @goldhaber2007 performs a similar analysis with the longitudinal data of North Carolina and comes to similar conclusions, strengthening the robustness of the results. Lastly, @goldhaber2015 examine the inequity in the distribution of teacher quality by high-needs groups in Washington state, and find that for all three measures of quality (teacher experience, licensure exam score, and value added), the distribution of teachers favors the less needy (as measured by free/reduced-price lunch status, minority status, and low prior academic achievement).

[^VA]: The most commonly cited expositions on value added, its validity, and so on are probably @rivkin, an extensive exploration of the predictive powers of empirical Bayes VA measures; and @chettyI and @chettyII, the largest-scale study of long-term inferences based on value added.

The aforementioned papers have tended to keep the collective bargaining aspect of salary determination for teachers out of the spotlight, if largely for reasons of data restrictions. Nevertheless, it stands to reason to believe that the rigid structure of union-negotiated contracts could serve to contribute in a large way to teacher turnover. @ballou give much descriptive evidence of the shape of the wage-tenure profile, rooted in a data set collected by the Department of Defense and published by the AFT. They find that seniority premia in education largely mirror those in more traditional white collar professins, that steeper profiles are associated with less turnover, and that district financial and demographic conditions alone are insufficient to explain variation in contracts. Another common (and recently quite controversial, as evidenced by the contention in the ongoing contract negotiations in Philadelphia) feature of union-negotiated teacher contracts are seniority priviliges--preferential treatments granted to teachers in voluntary and involuntary transfers. @moe codes contracts from 158 districts in California according to the strength of seniority rights therein guaranteed to teachers and finds that such rights are associated with the distribution of teachers across schools (measuring quality as experience and certification) in a way that serves to harm minorities. Revisiting California with a slightly different sample and definition of the “determinacy” of the contracts with respect to seniority, @koski come to the opposite conclusion--that there is no such relationship. As a rebuttal, @anzia pin the difference in results on the exclusion in @moe of small school districts, where it appears that the entrenchment of bureaucracy falters and the rigidity of contract language wane, a claim which they support by repeating their analysis with the inclusion of an interaction for district size--indeed, for small districts the result of @koski holds, while the insight of @moe holds in larger districts. @cohenvogel use data from Florida and their results align with those of @koski (though they neglect to nuance their results by district size).

Finally, an emerging but still immature strand of literature is beginning to look at the potential for transfer bonuses and retention incentives to positively affect student outcomes. @fulbeck analyzes a scheme in place in Denver whereby teachers who choose to transfer to high-needs schools (low-performing) are given recurring bonus pay, and those initially stationed there are given retention incentives. She concludes that recipients of incentives are significantly less likely to switch jobs, as driven by a reduction in district exit rates and especially by teachers whos incentive payments exceed \$5,000. @glazerman evaluate the Talent Transfer Initiative, a randomized controlled trial conducted in 10 districts whereby high-performance teachers were given \$20,000 over the course of two years as reward for transferring the identified high-needs schools, and conclude that there were significant effects on teacher retention as well as on student outcomes. 

# Data

# Results

Like @hanushek (HKR) found in Texas, most turnover in Wisconsin is happening within districts and out of the profession. In Wisconsin, the fraction of teachers transitioning among districts is vanishingly small after a "burn-in" period of roughly 6 years -- only `r round(100*teachers[total_exp_floor >= 7, mean(move_district_next)], 1)` of such teachers do so (compared with `r round((4.5*75284 + 2.5*165873 + 0.7*6978)/(75284+165873+6978), 1)`% for the comparable group in HKR), but is still relatively higher among the youngest teachers -- roughly twice as high for the "probationary" teachers (1-3 years' experience) as for teachers with 7-11 years' experience in both states. Movement patterns within districts in the two states are very similar, lending weight to teachers "earning their stripes" within a district to be able to choose the best schools.

```{r table_1, results = 'asis', message = FALSE}
library(funchir)
T1 = rbind(
  teachers[ , .(`Remain in Same School` = 100*sum(!move_school_next)/.N,
                `Change Schools Within District` = 
                  100*sum(move_within_next)/.N,
                `Switch Districts` = 100*sum(move_district_next)/.N,
                `Exit Wisconsin \\mbox{Public Schools}` = 100*sum(quit_next)/.N,
                `Number of Teachers` = prettyNum(.N, big.mark = ',')),
            keyby = .(`Teacher Experience` = exp_split)],
  teachers[ , .(`Teacher Experience` = 'All',
                `Remain in Same School` = 100*sum(!move_school_next)/.N,
                `Change Schools Within District` = 
                  100*sum(move_within_next)/.N,
                `Switch Districts` = 100*sum(move_district_next)/.N,
                `Exit Wisconsin \\mbox{Public Schools}` = 100*sum(quit_next)/.N,
                `Number of Teachers` = prettyNum(.N, big.mark = ','))])
suppressWarnings(
  tbl <- capture.output(print(xtable(
    T1, caption =
      'Year-to-year Transitions of Teachers by Experience, 2000-08', 
    label = 'tbl:move_by_exp', digits = 1L, 
    align = paste0('lp{.103\\linewidth}p{.14\\linewidth}',
                   'p{.15\\linewidth}p{.09\\linewidth}',
                   'p{.16\\linewidth}R{.16}')),
    sanitize.text.function = identity))
)
idx = grep('^Teacher', tbl) - 1L
tbl = c(tbl[1L:idx], 
        ' & \\multicolumn{4}{c}{Percent of Teachers Who} & \\\\ \\cline{2-5}',
        tbl[(idx+1L):length(tbl)])
cat(tbl, sep = '\n')
```

```{r table_2, results = 'asis'}
share_change = 
  #use 2006, not incl_rng, because of the artificial change in 
  #  shares induced by the change in how urbanicity is calculated
  teachers[year %in% c(2000, 2006), .N, by = .(year, urbanicity_s)
           ][ , shr := N/sum(N), by = year
              ][order(year), .(delta = diff(shr)), by = urbanicity_s]
T2 = rbind(
  teachers[ , {
    idx = move_district_next
    N_move = sum(idx)
    c(as.list(table2(urbanicity_s_next[idx], prop = TRUE, pct = TRUE)),
      list(n_change = prettyNum(N_move, big.mark = ','), 
           pct_origin = 100*N_move/.N,
           change_share = 
             share_change[.BY, sprintf('%.1f\\%%', 100*delta), on = 'urbanicity_s']))
  }, keyby = urbanicity_s],
  teachers[total_exp_floor <= 3, {
    idx = move_district_next
    N_move = sum(idx)
    c(as.list(table2(urbanicity_s_next[idx], prop = TRUE, pct = TRUE)),
      list(n_change = prettyNum(N_move, big.mark = ','), 
           pct_origin = 100*N_move/.N))
  }, keyby = urbanicity_s],
  fill = TRUE)

setnames(T2, grep('[^y]_', names(T2)), character(3L))
T2[ , c('Origin Community', 'urbanicity_s') :=
      .(paste0('\\quad ', urbanicity_s), NULL)]
setcolorder(T2, c(ncol(T2), seq_len(ncol(T2) - 1L)))

tbl = capture.output(print(xtable(
  T2, caption = paste('Destination Community Type for Teachers',
                      'Changing Districts, by Origin Community',
                      'Type and Teacher Experience Level'),
  label = 'tbl:markov', digits = 1L, align = 'llrrrrrrr'),
  sanitize.text.function = identity, floating.environment = 'sidewaystable'))
idx = grep('^Origin', tbl)
mrow = function(x, n) paste0('\\multirow{', n, '}{*}{', x, '}')
mcol = function(x, n) paste0('\\multicolumn{', n, '}{c}{', x, '}')
pbox = function(x, p) paste0('\\parbox{', p, '\\linewidth}{', x, '}')
tbl = c(tbl[1L:(idx - 1L)],
        paste0(paste('', mcol(mrow('Percent of Teachers Who Move to', 2), 4),
                     mrow(pbox('Number Teachers Changing Districts', .09), 4),
                     mrow(pbox('Percent of Origin Teachers', .07), 4),
                     mrow(pbox('Change in Share of Teachers 2000-06', .07), 4), 
                     sep = ' & '), '\\\\'),
        ' & \\multicolumn{4}{c}{} & & & \\\\ \\cline{2-5}',
        '& & & & & & & \\\\', tbl[idx + 0L:1L],
        'I. All teachers & & & & & & & \\\\', tbl[idx + (2L:5L)], 
        '\\multicolumn{3}{l}{II. Probationary teachers ' %+% 
          '(1-3 years experience)}' %+%
          ' & & & & & \\\\', tbl[(idx+6L):length(tbl)])
cat(tbl, sep = '\n')
```

```{r table_3, results = 'asis'}
T3 = 
  teachers[move_district_next & total_exp_floor <= 10, {
    dl = lapply(
      list(schedule_lsalary_next - schedule_lsalary_next_cf,
           schedule_lsalary_resid_next - schedule_lsalary_resid_next_cf,
           pct_prof_next_d - pct_prof_d, pct_hisp_next_d - pct_hisp_d,
           pct_black_next_d - pct_black_d, pct_frl_next_d - pct_frl_d),
      na.omit)
    .(var = paste0(rep(seq_len(6L), each = 2L), rep(c('M', 'S'), 6L)),
      val = c(rbind(sapply(dl, mean),
                    sapply(dl, function(x) sd(x)/sqrt(length(x))))))
  }, keyby = .(exp_split, gender)
  ][ , dcast(.SD, var ~ gender + exp_split, value.var = 'val')]
T3[ , all := 
      teachers[move_district_next & total_exp_floor <= 10, {
        dl = lapply(
          list(schedule_lsalary_next - schedule_lsalary_next_cf,
               schedule_lsalary_resid_next - schedule_lsalary_resid_next_cf,
               pct_prof_next_d - pct_prof_d, pct_hisp_next_d - pct_hisp_d,
               pct_black_next_d - pct_black_d, pct_frl_next_d - pct_frl_d),
          na.omit)
        c(rbind(sapply(dl, mean),
                sapply(dl, function(x) sd(x)/sqrt(length(x)))))}]]

T3[ , var := rep(c('Base year salary (log)', 'Adjusted salary (log)',
                   'Percent proficient', 'Percent Hispanic',
                   'Percent black', 'Percent subsidized lunch'), each = 2L)]
T3[ , type := rep(c('m', 's'), 6L)]
cols = setdiff(names(T3), c('var', 'type'))
sidx = T3$type == 's'
pidx = grepl('^Percent', T3$var)
T3[ , (cols) := lapply(.SD, function(x) {
  o = character(length(x))
  o[!pidx & !sidx] = sprintf('%.3f', x[!pidx & !sidx])
  o[!pidx & sidx] = sprintf('(%.3f)', x[!pidx & sidx])
  o[pidx & !sidx] = sprintf('%.1f%%', 100*x[pidx & !sidx])
  o[pidx & sidx] = sprintf('(%.1f%%)', 100*x[pidx & sidx])
  o}), .SDcols = cols]
T3[type == 's', var := '']
T3[ , type := NULL]
setnames(T3, c('', rep(exp_lab[1L:3L], 2L), ''))

tbl = capture.output(print(xtable(
  T3, caption = paste('Average Change in Salary and District Student',
                      'Characteristics (and Standard Deviations) for',
                      'Teachers Changing Districts, by Gender and Experience'),
  label = 'tbl:change_by_ge', align = 'llccccccc'), 
  floating.environment = 'sidewaystable'))

idx = grep('years', tbl)

tbl = c(tbl[1L:(idx - 1L)],
        paste0(paste('', mcol('Men by Experience Class', 3L),
                     mcol('Women by Experience Class', 3L),
                     mrow(pbox('All Teachers 0-9 Years', .09), 2L),
                     sep = ' & '), '\\\\ \\cline{2-4} \\cline{5-7}'),
        tbl[idx:length(tbl)])

cat(tbl, sep = '\n')
```

```{r table_4, results = 'asis'}
se = function(x) sd(x)/sqrt(length(x))
T4D = 
  teachers[move_district_next & urbanicity_next_d == 'Suburban' & 
             urbanicity_d %in% c('Large Urban', 'Suburban') & 
             total_exp_floor <= 10, {
               dl = lapply(
                 list(schedule_lsalary_next - schedule_lsalary_next_cf,
                      schedule_lsalary_resid_next -
                        schedule_lsalary_resid_next_cf,
                      pct_prof_next_d - pct_prof_d, 
                      pct_hisp_next_d - pct_hisp_d,
                      pct_black_next_d - pct_black_d, 
                      pct_frl_next_d - pct_frl_d),
                 na.omit)
               .(var = paste0(rep(seq_len(6L), each = 2L),
                              rep(c('M', 'S'), 6L)),
                 val = c(rbind(sapply(dl, mean), sapply(dl, se))))
             }, keyby = urbanicity_d
           ][ , dcast(.SD, var ~ urbanicity_d, value.var = 'val')]

T4S = 
  teachers[move_district_next & urbanicity_next_d == 'Suburban' & 
             urbanicity_d %in% c('Large Urban', 'Suburban') & 
             total_exp_floor <= 10, {
               dl = lapply(
                 list(pct_prof_next_s - pct_prof_s,
                      pct_hisp_next_s - pct_hisp_s,
                      pct_black_next_s - pct_black_s,
                      pct_frl_next_s - pct_frl_s),
                 na.omit)
               .(var = paste0(rep(3L:6L, each = 2L), rep(c('M', 'S'), 4L)),
                 val = c(rbind(sapply(dl, mean), sapply(dl, se))))
             }, keyby = urbanicity_d
           ][ , dcast(.SD, var ~ urbanicity_d, value.var = 'val')]

T4 = merge(T4D, T4S, by = 'var', all.x = TRUE)

T4[ , var := rep(c('Base year salary (log)', 'Adjusted salary (log)',
                   'Percent proficient', 'Percent Hispanic',
                   'Percent black', 'Percent subsidized lunch'), each = 2L)]
T4[ , type := rep(c('m', 's'), 6L)]
cols = setdiff(names(T4), c('var', 'type'))
sidx = T4$type == 's'
pidx = grepl('^Percent', T4$var)
T4[ , (cols) := lapply(.SD, function(x) {
  o = character(length(x))
  o[!pidx & !sidx] = sprintf('%.3f', x[!pidx & !sidx])
  o[!pidx & sidx] = sprintf('(%.3f)', x[!pidx & sidx])
  o[pidx & !sidx] = sprintf('%.1f\\%%', 100*x[pidx & !sidx])
  o[pidx & sidx] = sprintf('(%.1f\\%%)', 100*x[pidx & sidx])
  o[is.na(x) & !sidx] = '---'
  o[is.na(x) & sidx] = ''
  o}), .SDcols = cols]
T4[type == 's', var := '']
T4[ , type := NULL]
T4[grepl('^Percent', var), var := paste('\\quad', var)]

tbl = capture.output(print(xtable(
  T4, caption = paste('Average Change in Salary and in District and Campus',
                      'Student Characteristics (and Standard Deviations)',
                      'for Teachers with 1-10 Years of Experience Who',
                      'Change Districts, by Community Type of Origin',
                      'and Destination District'),
  label = 'tbl:change_by_urb', 
  align = paste0('llp{.12\\textwidth}p{.12\\textwidth}',
                 'p{.12\\textwidth}p{.12\\textwidth}')),
  sanitize.text.function = identity))

idx = grep('Urban', tbl) - 1L

tbl = c(tbl[1L:idx],
        paste0(paste('', mcol('District Average Characteristics', 2L),
                     mcol('Campus Average Characteristics', 2L),
                     sep = ' & '), '\\\\ \\cline{2-5}'),
        ' & Large Urban to Suburban & Suburban to Suburban & ' %+%
          'Large Urban to Suburban & Suburban to Suburban \\\\',
        tbl[(idx + 2L):length(tbl)])

idx = grep('quad', tbl)[1L] - 1L

tbl = c(tbl[1L:idx], 'Average Student Characteristics & & & & \\\\',
        tbl[(idx+1L):length(tbl)])

cat(tbl, sep = '\n')
```

```{r table_5, results = 'asis'}
T5D = 
  teachers[move_district_next & total_exp_floor <= 10 & 
             ethnicity_main %in% c('Black', 'Hispanic'), {
               dl = lapply(
                 list(pct_prof_next_s - pct_prof_s, 
                      pct_hisp_next_s - pct_hisp_s,
                      pct_black_next_s - pct_black_s, 
                      pct_frl_next_s - pct_frl_s),
                 na.omit)
               .(var = c(paste0(rep(1L:4L, each = 2L),
                                rep(c('M', 'S'), 4L)), 'N'),
                 val = c(rbind(sapply(dl, mean), sapply(dl, se)), .N))
             }, keyby = ethnicity_main
           ][ , dcast(.SD, var ~ ethnicity_main, value.var = 'val')]

T5S = 
  teachers[move_within_next & total_exp_floor <= 10 & 
             ethnicity_main %in% c('Black', 'Hispanic'), {
               dl = lapply(
                 list(pct_prof_next_s - pct_prof_s, 
                      pct_hisp_next_s - pct_hisp_s,
                      pct_black_next_s - pct_black_s, 
                      pct_frl_next_s - pct_frl_s),
                 na.omit)
               .(var = c(paste0(rep(1L:4L, each = 2L), 
                                rep(c('M', 'S'), 4L)), 'N'),
                 val = c(rbind(sapply(dl, mean),
                               sapply(dl, se)), .N))
             }, keyby = ethnicity_main
           ][ , dcast(.SD, var ~ ethnicity_main, value.var = 'val')]

T5 = merge(T5D, T5S, by = 'var', all.x = TRUE)

T5[ , var := rep(c('Percent proficient', 'Percent Hispanic',
                   'Percent black', 'Percent subsidized lunch',
                   'Number of teachers'), c(2L, 2L, 2L, 2L, 1L))]
T5[ , type := c(rep(c('m', 's'), 4L), 'n')]
cols = setdiff(names(T5), c('var', 'type'))
sidx = T5$type == 's'
pidx = grepl('^Percent', T5$var)
nidx = T5$type == 'n'
T5[ , (cols) := lapply(.SD, function(x) {
  o = character(length(x))
  o[pidx & !sidx] = sprintf('%.1f%%', 100*x[pidx & !sidx])
  o[pidx & sidx] = sprintf('(%.1f%%)', 100*x[pidx & sidx])
  o[nidx] = prettyNum(x[nidx], big.mark = ',')
  o}), .SDcols = cols]
T5[type == 's', var := '']
T5[ , type := NULL]

tbl = capture.output(print(xtable(
  T5, caption = paste('Average Change in District and Campus Student',
                      'Characteristics (and Standard Deviations)',
                      'for Black and Hispanic Teachers with 1-10',
                      'Years of Experience who Change Campuses'),
  label = 'tbl:change_by_eth', 
  align = paste0('llp{.1\\textwidth}p{.1\\textwidth}',
                 'p{.1\\textwidth}p{.1\\textwidth}'))))

idx = grep('Black.x', tbl)

tbl = c(tbl[1L:(idx - 1L)],
        paste0(paste('', mcol('Between District Moves', 2L),
                     mcol('Within District Moves', 2L),
                     sep = ' & '), '\\\\ \\cline{2-5}'),
        ' & Black Teachers & Hispanic Teachers & ' %+%
          'Black Teachers & Hispanic Teachers \\\\',
        tbl[(idx + 1L):length(tbl)])

cat(tbl, sep = '\n')
```

```{r table_6, results = 'asis'}
T6 = 
  melt(teachers, id.vars = c('year', 'move_district_next', 
                             'move_within_next', 'quit_next'), 
       measure.vars = paste0(q_var, '_q'), na.rm = TRUE, value.factor = TRUE
       )[ , .(`Probability Teachers Move to New School within District` = 
                mean(move_within_next),
              `Probability Teachers Move to New District` = 
                mean(move_district_next),
              `Probability Teachers Exit Public Schools` = 
                mean(quit_next)), 
          keyby = .(variable, value)]
T6[variable == 'lwage_resid_avg_q', grep('within', names(T6)) := NA]
T6[ , variable := NULL]
cols = grep('^Prob', names(T6))
T6[ , (cols) := lapply(.SD, function(x) {
  o = sprintf('%.1f\\%%', 100*x)
  is.na(o) = is.na(x); o}), .SDcols = cols]
T6[ , value := paste0('\\quad ', value)]
setnames(T6, 'value', 'Quartile of Distribution')

tbl = capture.output(print(xtable(
  T6, caption = paste('School Average Transition Rates by Distribution of',
                      'Residual Teacher Salary and Student Demographic',
                      'Characteristics (data weighted by number of',
                      'teachers in school)'),
  label = 'tbl:change_by_quartile',
  align = 'lp{.25\\textwidth}p{.13\\textwidth}p{.13\\textwidth}p{.13\\textwidth}'), 
  sanitize.text.function = identity, NA.string = '---'))

idx = grep('Highest', tbl)
tbl[idx] = paste0(c('Residual salary', 'Percent proficient',
                    'Percent eligible for reduced-price lunch',
                    'Percent Black', 'Percent Hispanic'),
                  ' & & & \\\\\n', tbl[idx])
cat(tbl, sep = '\n')
```

```{r table_7, results = 'asis'}
reg_l = 
  lapply(split(teachers, f = teachers$exp_split),
         function(DT) 
           DT[ , lm(leave_next ~
                      schedule_lbase_ba_salary*gender + 
                      pct_prof_s + pct_frl_s + 
                      ethnicity_main*pct_black_s + 
                      ethnicity_main*pct_hisp_s + factor(year) + 
                      urbanicity_s + total_exp_floor + 
                      I(total_exp_floor^2) + n_students_d + class_size_d)])

coefn = setNames(nm = names(coef(reg_l[[1L]])))
coefn[seq_len(length(coefn))] = 'x'
idx = match(c('schedule_lbase_ba_salary', 'schedule_lbase_ba_salary:genderF',
              'pct_prof_s', 'pct_frl_s', 'pct_black_s', 'pct_hisp_s',
              paste0('ethnicity_main', 
                     c('Black', 'Black', 'Hispanic', 'Hispanic'), ':pct_',
                     c('black', 'hisp', 'black', 'hisp'), '_s')), names(coefn))
coefn[idx] = 
  c('First year base salary (log)', 'First year base salary (log) * female',
    'Percent proficient', 'Percent eligible for subsidized lunch',
    'Percent Black', 'Percent Hispanic', 'Black * percent Black',
    'Black * percent Hispanic', 'Hispanic * percent Black',
    'Hispanic * percent Hispanic')
reg_l = lapply(reg_l, function(x) {
  names(x$coefficients) = coefn[names(coef(x))]; x})
tbl = capture.output(texreg(
  reg_l, omit.coef = '^x', label = 'tbl:reg_lpm', 
  caption = paste('Estimated Effects of Starting Teacher Salary and Student',
                  'Demographic Characteristics on the Probability that',
                  'Teachers Leave School Districts, by Experience (linear',
                  'probability models; standard errors in parentheses)'),
  groups = list('Campus average student characteristics' = 3:6,
                'Interactions' = 7:10), 
  reorder.coef = c(1, 6, 2:5, 7:10),
  include.rsquared = FALSE, include.adjrs = FALSE, include.rmse = FALSE))

idx = grep('years', tbl)
tbl = c(tbl[1L:(idx - 1L)],
        ' & \\multicolumn{4}{c}{Teacher Experience} \\\\ \\cline{2-6}',
        tbl[idx:length(tbl)])

idx = grep('^Num.*obs', tbl)
#print-formatting 1k+ numbers
tbl[idx] = gsub('(\\d{1,3})(?=\\d{3}+(?!\\d))', '\\1,', tbl[idx], perl = TRUE)
tbl[idx] = gsub('Num. obs.   ', 'Observations', tbl[idx], fixed = TRUE)

#exclude the extra blank rows introduced by texreg's group option
cat(tbl[grep('[^ &\\]', tbl)], sep = '\n')
```

```{r table_8, results = 'asis', cache = TRUE}
reg_l = 
  lapply(split(teachers, by = 'exp_split', sorted = TRUE),
         function(DT) DT[ , lm(leave_next ~
                                 schedule_lbase_ba_salary*gender + 
                                 pct_prof_s + pct_frl_s + 
                                 ethnicity_main*pct_black_s + 
                                 ethnicity_main*pct_hisp_s + factor(year) + 
                                 factor(district_fill) + total_exp_floor + 
                                 I(total_exp_floor^2) +
                                 n_students_d + class_size_d)])

#some districts are dropped by lm
#  (insufficient variation given low frequency of moves & other covariates)
coefn = setNames(nm = Reduce(union, lapply(reg_l, function(x) names(coef(x)))))
coefn[seq_len(length(coefn))] = 'x'
idx = match(c('schedule_lbase_ba_salary', 'schedule_lbase_ba_salary:genderF',
              'pct_prof_s', 'pct_frl_s', 'pct_black_s', 'pct_hisp_s',
              paste0('ethnicity_main', 
                     c('Black', 'Black', 'Hispanic', 'Hispanic'), ':pct_',
                     c('black', 'hisp', 'black', 'hisp'), '_s')), names(coefn))
coefn[idx] = 
  c('First year base salary (log)', 'First year base salary (log) * female',
    'Percent proficient', 'Percent eligible for subsidized lunch',
    'Percent Black', 'Percent Hispanic', 'Black * percent Black',
    'Black * percent Hispanic', 'Hispanic * percent Black',
    'Hispanic * percent Hispanic')
reg_l = lapply(reg_l, function(x) {
  names(x$coefficients) = coefn[names(coef(x))]; x})
tbl = capture.output(texreg(
  reg_l, omit.coef = '^x', label = 'tbl:reg_lpm_fe', 
  caption = paste('Estimated Effects of Starting Teacher Salary and Student',
                  'Demographic Characteristics on the Probability that',
                  'Teachers Leave School Districts with District Fixed',
                  'Effects, by Experience (linear probability models;',
                  'standard errors in parentheses)'),
  groups = list('Campus average student characteristics' = 3:6,
                'Interactions' = 7:10), 
  reorder.coef = c(1, 6, 2:5, 7:10),
  include.rsquared = FALSE, include.adjrs = FALSE, include.rmse = FALSE))

idx = grep('years', tbl)
tbl = c(tbl[1L:(idx - 1L)],
        ' & \\multicolumn{4}{c}{Teacher Experience} \\\\ \\cline{2-6}',
        tbl[idx:length(tbl)])
idx = grep('^Num.*obs', tbl)
#print-formatting 1k+ numbers
tbl[idx] = gsub('(\\d{1,3})(?=\\d{3}+(?!\\d))', '\\1,', tbl[idx], perl = TRUE)
tbl[idx] = gsub('Num. obs.   ', 'Observations', tbl[idx], fixed = TRUE)
#exclude the extra blank rows introduced by texreg's group option
cat(tbl[grep('[^ &\\]', tbl)], sep = '\n')
```

```{r table_9, eval = FALSE}
library(mlogit)
teachersML = 
  melt(teachers, id.vars =
         c('teacher_id', 'year', 'schedule_lbase_ba_salary', 
           'gender', 'pct_prof_s', 'pct_frl_s', 'urbanicity_s',
           'ethnicity_main', 'pct_black_s', 'pct_hisp_s',
           'total_exp_floor', 'n_students_d', 'class_size_d','exp_split'), 
       measure.vars = c('stay_next', 'move_district_next', 'quit_next'),
       variable.name = 'alternative', value.name = 'chosen')
setorder(teachersML, teacher_id, year, alternative)

teachersML[ , ID := .GRP, by = .(teacher_id, year)]

mnl_l = 
  lapply(split(teachersML, by = 'exp_split', sorted = TRUE), function(DT) {
    print(head(DT))
    setattr(DT, 'class', c('mlogit.data', class(DT)))
    setattr(DT, 'index', with(DT, data.table(alt = alternative, chid = ID)))
    mlogit(chosen ~ 0 | 
             schedule_lbase_ba_salary*gender + pct_prof_s + pct_frl_s + 
             ethnicity_main*pct_black_s + ethnicity_main*pct_hisp_s + 
             #need to include dummy for year, but solution fails
             urbanicity_s + total_exp_floor + 
             I(total_exp_floor^2) + n_students_d + class_size_d, data = DT)
  })
texreg(mnl_l)
```

# Conclusion

# References
